Model_CFU

### 2016-12-15 Projects/HumanCdGF_cdiff_colonization.Rmd
investigate bimodality of one mouse rsq

om_import_otus <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
colnames(om_import_otus) <- seq(ncol(om_import_otus))
om_otu_rank <- apply(om_import_otus, 2, function(x) rank(x, na.last = F))
om_otu_rank <- om_otu_rank*as.numeric(om_rsq[,3])
om_raw_rank <- sort(apply(om_otu_rank, 1, function(x) sum(x, na.rm = T)), decreasing = T)
med_otu_importance_om <- data.frame(otu = names(om_raw_rank),
                                 importance = om_raw_rank)

>>>>

all_otus_importance <- gather(om_import_otus,OTU, import_val,-otu)[,-2]
all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
                      filter(pos_otu, votes == 1) %>% select(Freq) - 
                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
              arrange(desc(Freq))
```

When using all OTUs the model preforms poorly. The median Rsq of the trainset was `r median(om_rsq$train_rsq)`, this is likely do to the features. This is a summary of the importance scores for all predictors:  
`r summary(unlist(select(om_import_otus, -otu)))`  
`r round(100*sum(om_import_otus[,-1]<=0, na.rm = T)/length(om_import_otus[,-1]<=0),2)`% of OTUs have an importance of 0 or less. The median Rsq of the test set was `r median(om_rsq$test_rsq)` and the Rsq of all predicted values was `r om_rsq_overall`.

This one mouse per source model is improved through reducing the amount of negative features, either utilizing the feature set via the top X % of predictors from the model of all mice or take only the features that are more often have an importance > 0 (OTUs : `r as.character(pos_otu$predictor)`).

```{r one mouse model - with OTUs positive in All OTU one mouse model}

#One per source - > 0
one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
    right_join(community_all_d0) %>% 
    filter(!is.na(log_cfu_med))) %>% 
    select(mouse_id, human_source, log_cfu_med,
           one_of(as.character(pos_otu$predictor)))
one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)

om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
om_pred_pos <- c()
om_rsq_pos <- c()
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_pos_otus_df %>% 
      group_by(human_source) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_pos_otus_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
                             importance = TRUE, ntree = n_tree, mtry = 10)
    om_pred_pos <- rbind(om_pred_pos,
                     data.frame(select(test, mouse_id, log_cfu_med),
                          predicted=predict(rf_om, test),
                          iter = i))
    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
                              mice = paste(train$mouse_id, collapse = ' '), 
                              train_rsq = tail(rf_om$rsq,1),
                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
      select(-IncNodePurity)
}

om_rsq_pos_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
           stringsAsFactors = F) %>% 
  gather(data_set, rsq, train, test)

om_rsq_pos_overall <- 1 - (sum((om_pred_pos$log_cfu_med - om_pred_pos$predicted)^2)/
               sum((om_pred_pos$log_cfu_med - mean(om_pred_pos$log_cfu_med))^2))

om_pred_pos <- rename(om_pred_pos, observed = log_cfu_med)

#all_otus_importance <- gather(om_import_pos_otus,OTU, import_val,-otu)[,-2]
#all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
#all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
#pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
#                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
#pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
#                      filter(pos_otu, votes == 1) %>% select(Freq) - 
#                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
#              arrange(desc(Freq))
```

One Mouse per Human Donor with OTUs > 0 give:
  *  median training set rsq of 0.08652421  
  *  median testing set rsq of 0.5331836  
  *  overall rsq of 0.5534567

One Mouse per Human Donor with the top 10% predictive OTUs when using all mice  
  *  median training set rsq of 0.07336261  
  *  median testing set rsq of 0.6070359  
  *  overall rsq of 0.633238

Looking at the % var explained values from each individual iteration, there appears to be a bimodal distribution in the performance.

```{r om_bimodality, fig.width= = 4, fig.height=4}
ggplot(om_rsq_pos_df, aes(rsq, color = data_set)) + 
  geom_density()
```

Using random forest to identify the mice best seperating the Rsq distribution

```{r one_mouse_rsq_predictors}

predict_rsq <- om_rsq_pos_df %>% 
                 filter(data_set == 'test')
predict_rsq$group_mean <- ifelse(predict_rsq$rsq > 0.62, T, F)

mice_diff <- data.frame(matrix(rep(NA, nrow(predict_rsq)*length(unique(colonized_cfu$mouse_id))),
                               nrow=nrow(predict_rsq)))
colnames(mice_diff) <- unique(colonized_cfu$mouse_id)
for(i in 1:nrow(mice_diff)){
    mice_diff[i,] <- ifelse(colnames(mice_diff) %in% strsplit(predict_rsq$mice, ' ')[[i]], 0,1)
}

predict_rsq_df <- cbind(rsq = predict_rsq$group_mean, mice_diff)
predict_rsq_df$rsq <- as.factor(predict_rsq_df$rsq)
set.seed(seed)
rsq_grp_rf <- randomForest(predict_rsq_df[,-1],predict_rsq_df$rsq, importance = T)
importance_rsq <- data.frame(mouse = dimnames(importance_rsq)[[1]],importance(rsq_grp_rf))
one_mouse_rsq_pred <- arrange(importance_rsq, desc(MeanDecreaseAccuracy)) %>% 
  select(mouse,MeanDecreaseAccuracy) %>% 
  mutate(Relative_MDA = MeanDecreaseAccuracy/max(MeanDecreaseAccuracy))

data.frame(mouse = factor(one_mouse_rsq_pred$mouse, 
                        levels = one_mouse_rsq_pred$mouse[order(one_mouse_rsq_pred$Relative_MDA)]), 
           Importance = one_mouse_rsq_pred$Relative_MDA) %>% 
ggplot(aes(x=mouse, y=Importance)) + 
   geom_point() + coord_flip() + 
    labs(y = 'Importance (Relative MDA)', x = NULL) + 
    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
          axis.ticks = element_line(size = 2),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          panel.grid.major.y = element_line(colour = "gray", linetype = "dotted"),
          panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
          axis.text.y = element_text(color = c(rep('black', length(one_mouse_rsq_pred$Relative_MDA)-
                                                     sum(one_mouse_rsq_pred$MeanDecreaseAccuracy > 0)),
                                               rep('red', sum(one_mouse_rsq_pred$MeanDecreaseAccuracy > 0))), 
                                     hjust = 0))

Due to small n?  
Due to incomplete source dependency?  
Ratio of train to test?
